SparseIDS: Learning Packet Sampling with Reinforcement Learning

### 2020-02-07 paper_rl/paper.tex
Tiny fix
work on related work

\glspl{rnn} have been shown to be valuable for constructing \glspl{ids} for network data. They allow determining if a flow is malicious or already before it is over, making it possible to take action immediately. However, considering the large number of packets that have to be inspected, the question of computational efficiency arises. We show that by using a novel \gls{rl}-based approach called \textit{\ours{}}, we can reduce the number of consumed packets by more than three fourths while keeping classification accuracy high. Comparing to various other sampling techniques, \ours{} consistently achieves higher classification accuracy by learning to sample only relevant packets. A major novelty of our \gls{rl}-based approach is that it can not only skip up to a predefined maximum number of samples like other approaches proposed in the domain of Natural Language Processing but can even skip arbitrarily many packets in one step, saving even more computational resources for long sequences. Inspecting \ours{}'s behavior of choosing packets shows that it adopts different sampling strategies for different attack types and network flows. Finally we build an automatic steering mechanism that can guide \ours{} in deployment to achieve a desired level of sparsity.

@@

\begin{enumerate}
%\item \textbf{choose optimally}, taking only the data that contain the most information and skip the less relevant ones.
% Fares:
\item \textbf{choose optimally}, taking only samples that contain the most information and skip the less relevant ones.
\item have a parameter that allows to \textbf{trade off} classification performance for sparsity (choosing fewer packets).
\item be \textbf{independent of the classifier} so that any classifier can be used and that the classifier doesn't have to be aware of the sampling strategy.
\item be \textbf{retrainable} so that the sampling can be continuously adapted depending on the current threat landscape.
\item be able to \textbf{skip arbitrarily} many packets since network flows can be very long but only the first couple of packets are needed to decide whether an attack occurred or not and the remaining packets 
\item be \textbf{lightweight}, so that computation efficiency is not lower because of the sampling overhead. 
\end{enumerate}

@@

There is a significant body of research on sampling for networks: \cite{zseby_sampling_2005} give an overview of common methods for packet sampling in IP networks. \cite{estan_building_2004} develop a scheme that adapts the sampling rate depending on available resources.

Several approaches have been proposed to implement sampling for \gls{ids}:  \cite{estan_new_2003} propose to single out large flows for \gls{ids} and sample fewer packets for increased computational efficiency. \cite{bakhoum_intrusion_2011} use an approach based on Markov chains to sample packets for \gls{ids}.
\cite{murali_kodialam_detecting_2003} lay out a game-theoretic model for determining which network paths are more vulnerable and require more packet sampling. 
\cite{ha_suspicious_2016} develop a system for Software Defined Networks, which aims to sample more packets from more vulnerable parts of the network.

Regarding \gls{rnn} for \gls{ids}, \cite{hartl_explainability_2019} build an LSTM-based recurrent classifier for Network Intrusion Detection and develop methods for explainability and to assess the security of recurrent classifiers.

Some works investigated ways to let \glspl{rnn} skip parts of sequences. The focus of these works was mostly Natural Language Processing and we know of no work that focuses on network traffic. The following papers provide an overview of this research domain:

\cite{yu_learning_2017} propose using \gls{rl} to let an LSTM learn to skim text. Their technique does not aim to maximize sparsity but only lets the \gls{rl} optimize classification performance for the sequence as a whole. Thus, if it learns to skip elements it is only because it is better for achieving good classification performance, not because it wants to increase sparsity.

\cite{campos_skip_2018} develop a technique that includes a \textit{skip gate} into LSTM and other recurrent cells. It works without \gls{rl} but on the downside, their method can only be trained once and not be adapted afterwards and also it explicitly depends on the the implementation of the underlying classifier, while we want a solution in which the sampling procedure and the classifier are independent.

\cite{seo_neural_2018} use an LSTM with one full state vector and one reduced one. At each step they decide whether to use the reduced or the full state vector and can reduce computational cost by using the reduced one. For network traffic we think that it is often not even necessary to have a reduced state vector since for some attack flows, already the first couple of packets might contain enough information and then the remaining packets of the flow do not have to be considered anymore at all.

@@

These proposed techniques focus their evaluation on text and hence do not fulfill all criteria we set out in \autoref{sec:introduction}. Furthermore, all techniques that use \gls{rl} use discrete actions, meaning that there is a hyperparameter $k$, which influences the maximum jump that is possible. However, for a network flow, already after the second packet it might be obvious that a flow is an attack and thus the remaining packets can be completely ignored. It is thus beneficial to not have a fixed maximum step size but a continuous one. This allows skipping arbitrarily many packets.

>>>>

\glspl{rnn} have been shown to be valuable for constructing \glspl{ids} for network data. They allow determining if a flow is malicious or not before it is over in a streaming fashion, making it possible to take action immediately. However, considering the large number of packets that have to be inspected, the question of computational efficiency arises. We show that by using a novel \gls{rl}-based approach called \textit{\ours{}}, we can reduce the number of consumed packets by more than three fourths while keeping classification accuracy high. Comparing to various other sampling techniques, \ours{} consistently achieves higher classification accuracy by learning to sample only relevant packets. A major novelty of our \gls{rl}-based approach is that it can not only skip up to a predefined maximum number of samples like other approaches proposed in the domain of Natural Language Processing but can even skip arbitrarily many packets in one step, saving even more computational resources for long sequences. Inspecting \ours{}'s behavior of choosing packets shows that it adopts different sampling strategies for different attack types and network flows. Finally we build an automatic steering mechanism that can guide \ours{} in deployment to achieve a desired level of sparsity.

@@

\begin{enumerate}
%\item \textbf{choose optimally}, taking only the data that contain the most information and skip the less relevant ones.
% Fares:
\item \textbf{choose optimally}, taking only samples that contain the most information and skip the less relevant ones.
\item have a parameter that allows to \textbf{trade off} classification performance for sparsity (choosing fewer packets).
\item be \textbf{independent of the classifier} so that any classifier can be used and that the classifier doesn't have to be aware of the sampling strategy.
\item be \textbf{retrainable} so that the sampling can be continuously adapted depending on the current threat landscape.
\item be able to \textbf{skip arbitrarily} many packets since network flows can be very long but only the first couple of packets are needed to decide whether an attack occurred or not and the remaining packets be lightweight, so that computation efficiency is not lower because of the sampling overhead. 
\end{enumerate}

@@

%There is a significant body of research on sampling for communication networks: \cite{zseby_sampling_2005} give an overview of common methods for packet sampling in IP networks. \cite{estan_building_2004} develop a scheme that adapts the sampling rate depending on available resources.

Sampling of network packets to minimize computational effort has been extensively studied and a significant body of reaserch has developed a variety of methods to cope with the complexity of traffic. In \cite{estan_building_2004}, authors have developed a scheme that varies the sampling rate depending on the available resources, making sampling framework more effective than conventional fixed-rate sampling. In view of the endless options of sampling techniques, the authors in \cite{zseby_sampling_2005} published an overview of common methods for packet sampling in IP networks.

%Several approaches have been proposed to implement sampling for \gls{ids}:  \cite{estan_new_2003} propose to single out large flows for \gls{ids} and sample fewer packets for increased computational efficiency. \cite{bakhoum_intrusion_2011} use an approach based on Markov chains to sample packets for \gls{ids}.
%\cite{murali_kodialam_detecting_2003} lay out a game-theoretic model for determining which network paths are more vulnerable and require more packet sampling. 
%\cite{ha_suspicious_2016} develop a system for Software Defined Networks, which aims to sample more packets from more vulnerable parts of the network.

As far as sampling for \glspl{ids} is concerned, several approaches have been proposed to sample packets retaining high detection performance. Authors in \cite{estan_new_2003} propose to single out large flows for \gls{ids} and sample fewer packets for increased computational efficiency. In \cite{bakhoum_intrusion_2011} an approach based on Markov chains to sample packets for \gls{ids} was used. \cite{murali_kodialam_detecting_2003} lay out a game-theoretic model for determining which network paths are more vulnerable and require more packet sampling and \cite{ha_suspicious_2016} develop a system for Software Defined Networks, which aims to sample more packets from more vulnerable parts of the network.

% Regarding \gls{rnn} for \gls{ids}, \cite{hartl_explainability_2019} build an LSTM-based recurrent classifier for Network Intrusion Detection and develop methods for explainability and to assess the security of recurrent classifiers.

Conversely, \gls{rnn}-based \glspl{ids} has been shown to perform very well. In \cite{hartl_explainability_2019}, authors build a LSTM-based recurrent classifier for Network Intrusion Detection and further develop explainability methods as well as recurrent-classifiers security assessment.

%Some works investigated ways to let \glspl{rnn} skip parts of sequences. The focus of these works was mostly Natural Language Processing and we know of no work that focuses on network traffic. The following papers provide an overview of this research domain:

In this context, some works investigated ways to let \glspl{rnn} skip parts of sequences. The focus of these works was mostly Natural Language Processing and we know of no work that focuses on network traffic. The following papers provide an overview of this research domain:

Authors in \cite{yu_learning_2017} propose the usage of \gls{rl} to make a LSTM network learn to skim text. Their technique does not aim to maximize sparsity but only lets the \gls{rl} optimize classification performance for the sequence as a whole. Thus, if it learns to skip elements it is only because it is better for achieving good classification performance, not because it wants to increase sparsity.

Another technique developed in \cite{campos_skip_2018} includes a \textit{skip gate} into LSTM and other recurrent cells. It works without \gls{rl} but on the downside, their method can only be trained once and not be adapted afterwards and also it explicitly depends on the the implementation of the underlying classifier, while we want a solution in which the sampling procedure and the classifier are independent.

A LSTM network with two state vectors (one full state vector and one reduced one) was used in \cite{seo_neural_2018}. At each step the network decides whether to use the reduced or the full state vector and can reduce computational cost by using the reduced one. For network traffic it is often not even necessary to have a reduced state vector since for some attack flows, already the first couple of packets might contain enough information and then the remaining packets of the flow do not have to be considered anymore at all \cite{hartl_explainability_2019}.

@@

These proposed techniques focus their evaluation on text and hence do not fulfill all criteria we set out in \autoref{sec:introduction}. Furthermore, all techniques that use \gls{rl} use discrete actions, meaning that there is a hyperparameter $k$, which influences the maximum jump that is possible. However, for a network flow, already after the second packet it might be obvious that a flow is an attack and thus the remaining packets can be completely ignored. It is therefore beneficial not to have a fixed maximum step size but a continuous one. This allows skipping arbitrarily many packets and gives more flexibility  to the sampling framework.
